---
title: "chooseR"
output:
    html_document:
        keep_md: yes
        toc_float: true
        toc: true
        code_folding: hide


---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      results='hide',
                      echo=FALSE)
```

## Activate conda env

Ensure getting all the packages in `requirements.txt` installed and running `activate conda ./env`

```{r packages_and_functions, include=FALSE}
library(tidyverse)
library(ggplot2)
library(Signac)
library(Seurat)
library(knitr)
library(Matrix)

# Set correct paths
source("/data/BSPC-Petros/multiome-mira/chooser/R/var.R")
```


## Load Seurat object

- **Input data** = RDS file storing a **seurat object** (before SCTransform & PCA)


```{r load_data}

# Read a QC-ed seuat object
obj <- readRDS(rds_path)

# Move active assay to Gene.Expression
# Satija's naming
if ("RNA" %in% names(obj)) {
    DefaultAssay(obj) <- "RNA"

    # Chris' naming
} else if ("Gene.Expression" %in% names(obj)) {
    DefaultAssay(obj) <- "Gene.Expression"
    obj <- RenameAssays(obj, Gene.Expression="RNA")
}




# Explore the output
obj



```

## Run SCTransform


```{r sctransform}

# Add pseudocount 0.01 to seurat's expression count matrix
# (avoids errors from missing or infinite values)
obj@assays$RNA@counts <- as(obj@assays$RNA@counts + 0.01,  # adds pseudocount
                            "dgCMatrix")                   # coerces to the original matrix type (dgCMatrix)


# Explore the output
kable(obj@assays$RNA@counts[1:5, 1:3])



# Run SCTransform
obj <- SCTransform(obj)


# Explore the matrix
obj

```

## Run PCA


```{r pca}



# Run PCA
obj <- RunPCA(obj)

```

## Set variables

```{r set_variables}

# Define the number of PCs to use, and which assay and reduction to use.
# We recommend testing a broad range of resolutions
# For more on picking the correct number of PCs, see:
# https://satijalab.org/seurat/v3.1/pbmc3k_tutorial.html
npcs <- obj@commands$RunPCA.SCT@params$npcs
assay <- "SCT"
reduction <- "pca"

# Create a directory for output plots
system(paste0("mkdir ", results_path))


```

## Run chooseR

- **Input = seurat object (after SCTransform and PCA)**


```{r chooser}

# Run pipeline
for (res in resolutions) {
  message(paste0("Clustering ", res, "..."))
  message("\tFinding ground truth...")

  # "Truths" will be stored at glue::glue("{reduction}.{assay}_res.{res}")
  obj <- find_clusters(obj,
                       npcs=npcs,
                       reduction = reduction,
                       assay = assay,
                       resolution = res
  )
  clusters <- obj[[glue::glue("{reduction}.{assay}_res.{res}")]]

  # Now perform iterative, sub-sampled clusters
  results <- multiple_cluster(obj,
                              n = nIter,
                              size = 0.8,
                              npcs = npcs,
                              res = res,
                              reduction = reduction,
                              assay = assay
  )

  # Now calculate the co-clustering frequencies
  message(paste0("Tallying ", res, "..."))
  # This is the more time efficient vectorisation
  # However, it exhausts vector memory for (nearly) all datasets
  # matches <- purrr::map(columns, find_matches, df = results)
  # matches <- purrr::reduce(matches, `+`)
  columns <- colnames(dplyr::select(results, -cell))
  mtchs <- matrix(0, nrow = dim(results)[1], ncol = dim(results)[1])
  i <- 1 # Counter
  for (col in columns) {
    message(paste0("\tRound ", i, "..."))
    mtchs <- Reduce("+", list(
      mtchs,
      find_matches(col, df = results)
    ))
    i <- i + 1
  }

  message(paste0("Scoring ", res, "..."))
  mtchs <- dplyr::mutate_all(
    dplyr::as_tibble(mtchs),
    function(x) dplyr::if_else(Re(x) > 0, percent_match(x), 0)
  )

  # Now calculate silhouette scores
  message(paste0("Silhouette ", res, "..."))
  sil <- cluster::silhouette(
    x = as.numeric(as.character(unlist(clusters))),
    dmatrix = (1 - as.matrix(mtchs))
  )
  saveRDS(sil, paste0(results_path, "silhouette_", res, ".rds"))

  # Finally, calculate grouped metrics
  message(paste0("Grouping ", res, "..."))
  grp <- group_scores(mtchs, unlist(clusters))
  saveRDS(grp, paste0(results_path, "frequency_grouped_", res, ".rds"))
  sil <- group_sil(sil, res)
  saveRDS(sil, paste0(results_path, "silhouette_grouped_", res, ".rds"))
}


```

## Plot {.tabset .tabset-pills}

```{r prep_plot}


# Create silhouette plot
# Read in scores and calculate CIs
scores <- purrr::map(
  paste0(results_path, "silhouette_grouped_", resolutions, ".rds"),
  readRDS
)




scores <- dplyr::bind_rows(scores) %>%
  dplyr::group_by(res) %>%
  dplyr::mutate("n_clusters" = dplyr::n()) %>%
  dplyr::ungroup()
meds <- scores %>%
  dplyr::group_by(res) %>%
  dplyr::summarise(
    "boot" = list(boot_median(avg_sil)),
    "n_clusters" = mean(n_clusters)
  ) %>%
  tidyr::unnest_wider(boot)

writexl::write_xlsx(meds, paste0(results_path, "median_ci.xlsx"))

# Find thresholds
threshold <- max(meds$low_med)
choice <- as.character(
  meds %>%
  dplyr::filter(med >= threshold) %>%
  dplyr::arrange(n_clusters) %>%
  tail(n = 1) %>%
  dplyr::pull(res)
)


```

### Silhouette distribution

```{r plot, results='asis'}

# And plot!
ggplot(meds, aes(factor(res), med)) +
  geom_crossbar(
    aes(ymin = low_med, ymax = high_med),
    fill = "grey",
    size = 0.25
  ) +
  geom_hline(aes(yintercept = threshold), colour = "blue") +
  geom_vline(aes(xintercept = choice), colour = "red") +
  geom_jitter(
    data = scores,
    aes(factor(res), avg_sil),
    size = 0.35,
    width = 0.15
  ) +
  scale_x_discrete("Resolution") +
  scale_y_continuous(
    "Silhouette Score",
    expand = c(0, 0),
    limits = c(-1, 1),
    breaks = seq(-1, 1, 0.25),
    oob = scales::squish
  ) +
  cowplot::theme_minimal_hgrid() +
  theme(
    axis.title = element_text(size = 8),
    axis.text = element_text(size = 7),
    axis.line.x = element_line(colour = "black"),
    axis.line.y = element_line(colour = "black"),
    axis.ticks = element_line(colour = "black"),
  )

ggsave(
  filename = paste0(results_path, "silhouette_distribution_plot.png"),
  dpi = 300,
  height = 3.5,
  width = 3.5,
  units = "in"
)

```

### Dot plot

```{r dot_plot, results='asis'}

# Finally, a dot plot of silhouette scores to help identify less robust clusters
# The initial pipe is to order the clusters by silhouette score
scores %>%
  dplyr::filter(res == choice) %>%
  dplyr::arrange(dplyr::desc(avg_sil)) %>%
  dplyr::mutate_at("cluster", ordered, levels = .$cluster) %>%
  ggplot(aes(factor(cluster), avg_sil)) +
    geom_point() +
    scale_x_discrete("Cluster") +
    scale_y_continuous(
      "Silhouette Score",
      expand = c(0, 0),
      limits = c(-1, 1),
      breaks = seq(-1, 1, 0.25),
      oob = scales::squish
    ) +
    cowplot::theme_minimal_grid() +
    theme(
      axis.title = element_text(size = 8),
      axis.text = element_text(size = 7),
      axis.line.x = element_line(colour = "black"),
      axis.line.y = element_line(colour = "black"),
      axis.ticks = element_line(colour = "black"),
    )

ggsave(
  filename = paste0(results_path, "silhouette_point_plot_", choice, ".png"),
  dpi = 300,
  height = 3.5,
  width = 3.5,
  units = "in"
)


```

## Save the clustered seurat object

```{r save}

# Save original data, with ground truth labels
saveRDS(obj, paste0(results_path, "clustered_data.rds"))
cat('Done saving seurat object.')

```

## Co-clustering {.tabset, tabset-pills}

```{r coclustering}

# First is a cluster average co-clustering heatmap
# Read the data
grp <- readRDS(paste0(results_path, "frequency_grouped_", choice, ".rds"))

# As the data is symmetrical, we do not need the upper triangle
grp <- grp %>%
  pivot_wider(names_from = "cell_2", values_from = "avg_percent") %>%
  select(str_sort(colnames(.), numeric = T)) %>%
  column_to_rownames("cell_1")
grp[lower.tri(grp)] <- NA
grp <- grp %>%
  as_tibble(rownames = "cell_1") %>%
  pivot_longer(-cell_1, names_to = "cell_2", values_to = "avg_percent") %>%
  mutate_at("cell_2", ordered, levels = unique(.$cell_1)) %>%
  mutate_at("cell_1", ordered, levels = unique(.$cell_1))

```

### Heatmap

```{r heatmap, results='asis'}

# And plot!
plot <- ggplot(grp, aes(factor(cell_1), cell_2, fill = avg_percent)) +
  geom_tile() +
  scale_x_discrete("Cluster", expand = c(0, 0)) +
  scale_y_discrete(
    "Cluster",
    limits = rev(levels(grp$cell_2)),
    expand = c(0, 0)
  ) +
  scale_fill_distiller(
    " ",
    limits = c(0, 1),
    breaks = c(0, 0.5, 1),
    palette = "RdYlBu",
    na.value = "white"
  ) +
  coord_fixed() +
  theme(
    axis.ticks = element_line(colour = "black"),
    axis.text = element_text(size = 6),
    axis.title = element_text(size = 8),
    legend.text = element_text(size = 7),
    legend.position = c(0.9, 0.9)
  ) +
  guides(fill = guide_colorbar(barheight = 3, barwidth = 1))

ggsave(
  plot = plot,
  filename = paste0(results_path, "coclustering_heatmap_", choice, ".png"),
  dpi = 300,
  height = 3.5,
  width = 3.5,
  units = "in"
)

```

## Silhouette score {.tabset .tabset-pills}

```{r silhouette_score}

# Let's add the silhouette scores to the Seurat object!
sil_scores <- readRDS(paste0(results_path, "silhouette_", choice, ".rds"))
sil_scores <- as.data.frame(sil_scores[, 3], row.names = Seurat::Cells(obj))
colnames(sil_scores) <- c("sil_score")
obj <- AddMetaData(obj, metadata = sil_scores)

# Let's visualise the selected cluster
# If your data has known  clusters, you could also visualise those!
# Remember, truths are in "glue({reduction}.{assay}_res.{choice})"
# Seurat Changes color scheme if you order your data, so we provide
# the following helper function to restore defaults
gg_color <- function(n) {
  hues <- seq(15, 375, length = n + 1)
  colours <- hcl(h = hues, c = 100, l = 65)[1:n]
  return(colours)
}


```

### UMAP

```{r umap, results='asis'}

plot <- DimPlot(
  obj,
  reduction = "umap",
  group.by = glue::glue("{reduction}.{assay}_res.{choice}"),
  pt.size = 0.5,
  # cols = gg_color(6) # Only necessary if you have ordered your clusters
)

ggsave(
  plot = plot,
  filename = paste0(results_path, choice, "_cluster_umap.png"),
  dpi = 300,
  height = 5,
  width = 5,
  units = "in"
)

plot

```

### UMAP with the silouette scores

```{r umap_silscores, results='asis'}

# We also find it useful to visualise the silhouette scores on the UMAP!
plot <- FeaturePlot(
  obj,
  "sil_score",
  reduction = "umap",
  pt.size = 0.5,
  min.cutoff = -1,
  max.cutoff = 1
) +
  scale_colour_distiller(
    palette = "RdYlBu",
    labels = c(-1, 0, 1),
    breaks = c(-1, 0, 1),
    limits = c(-1, 1)
  )


plot

ggsave(
  plot = plot,
  filename = paste0(results_path, choice, "_silhouette_umap.png"),
  dpi = 300,
  height = 5,
  width = 5,
  units = "in"
)

```

### Cluster heatmap

```{r cluster_heatmap, results='asis'}

# This dataset does have annotated cell types, so we can create a heatmap
# to see how well they line up. Remember, "truths" will always be in
# glue::glue("{reduction}.{assay}_res.{choice}"), but annotated cell types
# might be in different locations. Be sure to change the names as needed.
ids <- as_tibble(
    obj[[c(glue::glue("{reduction}.{assay}_res.{choice}"), "CellType")]]
  ) %>%
  mutate_at(
    c(glue::glue("{reduction}.{assay}_res.{choice}"), "CellType"),
    as.factor
  ) %>%
  group_by(pca.SCT_res.1.6, CellType) %>%
  summarise("count" = n()) %>%
  ungroup() %>%
  complete(pca.SCT_res.1.6, CellType, fill = list("count" = 0)) %>%
  # Count suggested clusters
  group_by(pca.SCT_res.1.6) %>%
  mutate("n_suggested" = sum(count)) %>%
  ungroup() %>%
  # Count known clusters
  group_by(CellType) %>%
  mutate("n_known" = sum(count)) %>%
  ungroup() %>%
  # Calculate statistics
  mutate("dice" = (2 * count) / (n_suggested + n_known)) %>%
  mutate(
    "pca.SCT_res.1.6" = fct_reorder(pca.SCT_res.1.6, dice, max, .desc = TRUE),
    "CellType" = fct_reorder(CellType, dice, max, .desc = TRUE)
  )

plot <- ggplot(ids, aes(pca.SCT_res.1.6, CellType, fill = dice)) +
  geom_tile() +
  scale_x_discrete("Suggested Clusters", expand = c(0, 0)) +
  scale_y_discrete(
    "Known Clusters",
    limits = rev(levels(ids$CellType)),
    expand = c(0, 0)
  ) +
  coord_fixed() +
  scale_fill_distiller(
    NULL,
    limits = c(0, 1),
    breaks = c(0, 0.5, 1),
    palette = "RdYlBu"
  ) +
  theme(
    axis.ticks = element_line(colour = "black"),
    axis.text = element_text(size = 6),
    axis.title = element_text(size = 8),
    legend.text = element_text(size = 7),
  ) +
  guides(fill = guide_colorbar(barheight = 3, barwidth = 0.5))


plot

ggsave(
  plot = plot,
  filename = paste0(results_path, "cluster_relation_heatmap.png"),
  dpi = 300,
  height = 3.5,
  width = 3.5,
  units = "in"
)

```

## Session info

```{r session_info}

sessionInfo()

```

